---
title: 'Core panel manuscript'
author: "Kane Foster"
date: "2025-02-03"
output: html_document
---

## Setup

```{r}
library(tidyverse)
library(ggpubr)
library(RColorBrewer)
library(viridis)
source('../resources/aes.R')
range01 <- function(x){(x-min(x))/(max(x)-min(x))}
load(file = '../preprocessing/data/comp.RData')
load(file = '../preprocessing/data/ords.RData')
obs = read.csv('../preprocessing/data/Tcell_obs.csv')
donor_md = read.csv('../preprocessing/data/import/metadata-donor.csv')
sort_id = read.csv('../resources/Foster_2024_sort_id.csv')
dir.create('../figures/flow')

library(lubridate)
library(survival)
library(ggsurvfit)
library(tidycmprsk)
library(gtsummary)
library(condsurv)
#library(maxstat)

# Data
import = read_csv('/Volumes/groupfolders/CI_Haematology_Yong/Kane Foster/manuscript2025-core_panel/data_merged.csv')
import$CD127n.pct_CD8 = 100-import$CD127p.pct_CD8
# Treg pct NonTum thr
#dat = import %>% filter(timepoint=='Baseline', !is.na(event), !is.na(treg.pct_nonTum), days < 3*365)
#mtHL = maxstat.test(Surv(days, event) ~ treg.pct_nonTum, data=dat, smethod="LogRank", pmethod="HL")
#mtHL$estimate #0.5288 
treg.pctNon_thr = 0.5288  
# CD127nCD8 thr
#dat = import %>% filter(timepoint=='Baseline',!is.na(event), !is.na(CD127n.pct_CD8), days < 3*365)
#mtHL = maxstat.test(Surv(days, event) ~ CD127n.pct_CD8, data=dat, smethod="LogRank", pmethod="HL")
#mtHL$estimate #40.6
CD127n.pct_thr = 40.6
# Treg pct CD4+
#dat = import %>% filter(timepoint=='Baseline', !is.na(event), !is.na(treg.pct_CD4), days < 3*365)
#mtHL = maxstat.test(Surv(days, event) ~ treg.pct_CD4, data=dat, smethod="LogRank", pmethod="HL")
#mtHL$estimate #5.64
treg.pctCD4_thr = 5.64

```


## scRNAseq CD127- % CD8+ vs PC1

```{r fig.height=2.5, fig.width=3.5}
exp = obs %>% select(obs_names,sample_id) %>% 
  filter(sample_id %in% ords$Tcell$PCA$sample_id) %>% 
  left_join(read.csv('../preprocessing/data/Tcell_exp.csv') %>% 
              select(obs_names,CD8A,IL7R)) %>% drop_na()

dat = exp %>% 
  filter(CD8A > 0) %>% mutate(gr = IL7R>0) %>% 
  group_by(sample_id) %>% summarise(pct = sum(gr)/n()*100) %>% 
  left_join(ords$Tcell$PCA) %>% 
  left_join(comp$Tcell %>% select(sample_id,cohort) %>% distinct())

dat %>% 
  mutate(cohort = factor(cohort, c('Non','MGUS','SMM','MM'))) %>% 
  ggplot(aes(PC1 * -1, 100 - pct))+
  geom_point(aes(color=cohort))+scale_color_manual(values=diagnosis_col)+
  geom_smooth(method='lm')+stat_cor()+
  labs(x = 'T-cell skewing', y='IL7R– % CD8+')-> pl
pl
ggsave(plot=pl,filename = '../figures/flow/cor-scRNA_CD127.PC1.pdf',width=3.5,height=2.5, dpi="retina")

```

## scRNAseq CD127-/+ expression

```{r fig.height=3, fig.width=3.75}

exp = read.csv('../preprocessing/data/Tcell_exp.csv') %>% 
  select(obs_names,CD8A,IL7R,SELL,CCR7,CD28,GZMB,GNLY,CX3CR1)
  
dat = exp %>% filter(CD8A > 0) %>% mutate(gr = IL7R>0) %>% 
  left_join(obs %>% select(obs_names,sample_id)) %>% 
  select(-CD8A,-IL7R,-obs_names) %>% 
  group_by(sample_id,gr) %>% summarise_all(.funs = mean)

dat %>% pivot_longer(!c(sample_id:gr)) %>% 
  group_by(name) %>% mutate(value=range01(value)) %>% 
  mutate(name = factor(name, c('SELL','CCR7','CD28','GZMB','GNLY','CX3CR1'))) %>% 
  mutate(gr=factor(gr,c('TRUE','FALSE'))) %>% 
  ggplot(aes(gr,value,fill=gr))+facet_wrap('name',ncol=3)+
  geom_violin()+geom_boxplot(outlier.alpha=0)+
  ylim(c(0,1.3))+
  ggpubr::stat_compare_means(comparisons = list(c('FALSE','TRUE')),
    method = 'wilcox.test',size=3.5,label.y = 1.1)+
  scale_fill_manual(values=c('FALSE'='dodgerblue','TRUE'='grey80'))-> pl
pl
ggsave(plot=pl,filename = '../figures/flow/cor-scRNA_CD127.markers.pdf',width=3.5,height=2.5, dpi="retina")
  
```


## CD127- % CD8+ vs PP, with age

```{r fig.height=2.5, fig.width=3.5}

dat = import %>% 
  filter(timepoint=='Baseline',!is.na(Entry_LB_paraprotein),!is.na(CD127p.pct_CD8)) 

print(dim(dat)[1])

dat %>% 
  ggplot(aes(Entry_LB_paraprotein,100 - CD127p.pct_CD8))+
  geom_point(aes(color=cohort))+scale_color_manual(values=diagnosis_col)+
  geom_smooth(method='lm',color='black')+stat_cor()+
  labs(y='IL7R– % CD8+')-> pl
pl
ggsave(plot=pl,filename = '../figures/flow/cor-CD127.PP.pdf',width=3.5,height=2.5, dpi="retina")

summary(lm(CD127p.pct_CD8 ~ Entry_LB_paraprotein + age, dat))

summary(lm(CD127p.pct_CD3 ~ Entry_LB_paraprotein + age, dat))
cor.test(100 - dat$CD127p.pct_CD3, dat$Entry_LB_paraprotein)
 
```


## CyTOF TemTerm vs PP

```{r fig.height=2.5, fig.width=3.5}


dat = read_csv('/Volumes/ritd-ag-project-rd017s-klyon56/Atlas_analysis_code/R-resub/data/CyTOF/CyTOF_reGateCGK-Proportions.csv') %>% left_join(read_csv('/Volumes/ritd-ag-project-rd017s-klyon56/Atlas_analysis_code/R-resub/data/CyTOF/CyTOF_reGateCGK-md.csv')) 




#var              all     preCOSMOS
#CD57p_GZMBp_CD8  P=0.84  P=0.29
#Tem_pctCD8       P=0.14  P=0.05
#Temra_pctCD8     P=0.56  P=0.2
#CD57p_GZMBp_CD8  P=0.74  P=0.2

dat %>% 
  ggplot(aes(paraprotein,Tem_pctCD8))+
  geom_point(aes(color=cohort))+scale_color_manual(values=diagnosis_col)+
  geom_smooth(method='lm',color='black')+
  ylim(c(5,40))+
  stat_cor()-> pl
pl
ggsave(plot=pl,filename = '../figures/flow/cor-CD8_Tem.PP.pdf',width=3.5,height=2.5, dpi="retina")
 
```




## COSMOS TTP overview

cohort
```{r fig.height=3, fig.width=3.5}
dat = import %>% filter(timepoint=='Baseline',!is.na(event),days < 3*365)

median(dat$days) / 365
range(dat$days) / 365

dim(dat)[1]
table(dat$event)
table(dat$event)/dim(dat)[1]
```

overall curve
```{r fig.height=3, fig.width=3.5}
dat = import %>% filter(timepoint=='Baseline',!is.na(event),days < 3*365)

survfit(Surv(days, event) ~ 1, data = dat) %>% 
  ggsurvfit(type = "risk", xscale='d_m')+
  add_confidence_interval()+
  add_censor_mark(size=1)+
  scale_ggsurvfit(x_scales = list( 
    breaks = seq(0,365*3.5,365/2), labels=seq(0,42,6), limits=c(0,365*3) 
  ))+
  add_risktable()+
  scale_y_continuous(breaks=seq(0,1,0.2),labels=seq(0,100,20),limits=c(0,1))+
  labs(x='Months',y='% progressed')+
  theme_classic()+
  theme(legend.position='bottom')->pl

ggsave(plot=print(pl, newpage = FALSE),filename = '../figures/flow/surv-all.pdf',width=3.5,height=3, dpi="retina")
```

key features associations
```{r fig.height=3, fig.width=3.5}
print(survdiff(Surv(days, event) ~ tumour.pct_live, data = dat)$pvalue)
print(survdiff(Surv(days, event) ~ Entry_LB_paraprotein, data = dat)$pvalue)
print(survdiff(Surv(days, event) ~ Entry_LB_FLC_ratio, data = dat)$pvalue)
```



## SMM risk and TTP, forest plot

Mayo
```{r fig.height=1.5, fig.width=4.5}

dat = import %>% filter(timepoint=='Baseline',!is.na(event)) %>% 
  mutate(Entry_Mayo_=ifelse(Entry_Mayo %in% c('high','UHR'),'high/UHR',Entry_Mayo)) %>% 
  mutate(Entry_Mayo_=factor(Entry_Mayo_,c('MGUS','low','intermediate','high/UHR'))) 

cox_model = coxph(Surv(days, event) ~ Entry_Mayo_, data = dat)
cox_summary <- summary(cox_model)
#Extract hazard ratios, confidence intervals, and p-values
hazard_ratios <- exp(coef(cox_model))
conf_int <- exp(confint(cox_model))
p_values <- cox_summary$coefficients[,5]
#Plot
dat.pl = data.frame(
  Var1 = rownames(cox_summary$conf.int),
  pvals=p_values,
  hr = hazard_ratios,
  lower = conf_int[,1],
  upper = conf_int[,2]
) %>% 
  left_join(data.frame(table(dat$Entry_Mayo_)) %>% mutate(Var1=paste0('Entry_Mayo_',Var1)))

dat.pl %>% 
  ggplot(aes(y = paste0(Var1,' n=',Freq), x = hr)) +
  geom_point(size = 3, color = "royalblue") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  geom_text(aes(label=round(pvals,2)),color='red')+
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10(breaks = c(0.1, 1, 10)) +
  labs(
    title = paste0('vs MGUS n=',table(dat$Entry_Mayo_)[1]),
    x = "Hazard ratio (95% CI)",
    y = "Variable"
  )+
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12)
  )-> pl
pl
ggsave(plot=pl,filename = '../figures/flow/forest-all_Mayo.PP.pdf',width=4.5,height=1.5, dpi="retina")

```

Entry_IMWG 20/2/20
```{r fig.height=1.5, fig.width=4.5}

dat = import %>% filter(timepoint=='Baseline',!is.na(event)) %>% 
  mutate(Entry_IMWG_=Entry_IMWG) %>% 
  mutate(Entry_IMWG_=ifelse(Entry_IMWG_ %in% c('high','UHR'),'high/UHR',Entry_IMWG_)) %>% 
  mutate(Entry_IMWG_=ifelse(Entry_IMWG_=='low-intermediate','low',Entry_IMWG_)) %>% 
  mutate(Entry_IMWG_=factor(Entry_IMWG_,c('MGUS','low','intermediate','high/UHR'))) 

cox_model = coxph(Surv(days, event) ~ Entry_IMWG_, data = dat)
cox_summary <- summary(cox_model)
#Extract hazard ratios, confidence intervals, and p-values
hazard_ratios <- exp(coef(cox_model))
conf_int <- exp(confint(cox_model))
p_values <- cox_summary$coefficients[,5]
#Plot
dat.pl = data.frame(
  Var1 = rownames(cox_summary$conf.int),
  pvals=p_values,
  hr = hazard_ratios,
  lower = conf_int[,1],
  upper = conf_int[,2]
) %>% 
  left_join(data.frame(table(dat$Entry_IMWG_)) %>% mutate(Var1=paste0('Entry_IMWG_',Var1)))

dat.pl %>% 
  ggplot(aes(y = paste0(Var1,' n=',Freq), x = hr)) +
  geom_point(size = 3, color = "royalblue") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  geom_text(aes(label=round(pvals,2)),color='red')+
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10(breaks = c(0.1, 1, 10)) +
  labs(
    title = paste0('vs MGUS n=',table(dat$Entry_IMWG_)[1]),
    x = "Hazard ratio (95% CI)",
    y = "Variable"
  )+
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12)
  )-> pl
pl
ggsave(plot=pl,filename = '../figures/flow/forest-all_IMWG.PP.pdf',width=4.5,height=1.5, dpi="retina")


```


## CyTOF TemTerm SMM/MM

```{r fig.width=1.25, fig.height=3} 

dat = read_csv('/Volumes/ritd-ag-project-rd017s-klyon56/Atlas_analysis_code/R-resub/data/CyTOF/CyTOF_reGateCGK-Proportions.csv') %>% left_join(read_csv('/Volumes/ritd-ag-project-rd017s-klyon56/Atlas_analysis_code/R-resub/data/CyTOF/CyTOF_reGateCGK-md.csv'))

#CD127n_pctCD8,Tem_pctCD8,Temra_pctCD8,CD57p_GZMBp_CD8

dat %>% 
  mutate(cohort=factor(cohort,c('SMM','MM'))) %>% 
  ggplot(aes(cohort,Tem_pctCD8,fill=cohort))+
  
  geom_boxplot(outlier.alpha = 0)+
  geom_jitter(width=0.2,height=0,size=0.2)+
  ggpubr::stat_compare_means(comparisons = list(c('SMM','MM')), method = 'wilcox.test',size=3.5)+
  scale_fill_manual(values=diagnosis_col)+
  #scale_y_log10()+
  #scale_y_continuous(breaks=c(0,0.25,0.5),limits=c(0,0.75),labels=c('0','0.25','0.5'))+
  labs(title='')+

  theme_classic()+
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size=12),
    axis.title.x = element_blank(), 
    axis.text.x = element_text(angle = 45, vjust = 1, hjust=1, size=11),
    axis.title.y = element_text(size=13), 
    axis.text.y = element_text(size=11)
  )->pl
pl
ggsave(plot=pl,filename = '../figures/flow/box-CyTOF_TemBM.pdf',width=1.25,height=3, dpi="retina")

```



## flow_CD127- % SMM-prg/non

```{r fig.width=1.65, fig.height=3} 

dat = import %>% filter(timepoint=='Baseline',!is.na(event), !is.na(CD127p.pct_CD8), days<(365*3))

dat$cohort_prg = dat$cohort
dat$cohort_prg[dat$event==1  ]='SMM_PRG'

dat %>% 
  ggplot(aes(cohort_prg,CD127n.pct_CD8,fill=cohort_prg))+
  
  geom_boxplot(outlier.alpha = 0)+
  geom_jitter(width=0.2,height=0,size=0.2)+
  ggpubr::stat_compare_means(comparisons =
    list(c('MGUS','SMM'),c('MGUS','SMM_PRG'),c('SMM','SMM_PRG')),
  method = 'wilcox.test',size=3.5)+
  scale_fill_manual(values=diagnosis_col)+
  #scale_y_log10()+
  #scale_y_continuous(breaks=c(0,0.25,0.5),limits=c(0,0.75),labels=c('0','0.25','0.5'))+
  geom_hline(yintercept = CD127n.pct_thr, color='blue')+
  labs(y='IL7R– % CD8+',title='')+

  theme_classic()+
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size=12),
    axis.title.x = element_blank(), 
    axis.text.x = element_text(angle = 45, vjust = 1, hjust=1, size=11),
    axis.title.y = element_text(size=13), 
    axis.text.y = element_text(size=11)
  )->pl
pl
ggsave(plot=pl,filename = '../figures/flow/box-flow_IL7Rn.pdf',width=1.65,height=3, dpi="retina")

```




## CD127- TTP


```{r fig.height=4, fig.width=3.5}

dat = import %>% 
  filter(timepoint=='Baseline',!is.na(event), !is.na(CD127p.pct_CD8), days<(3*365))

dat$hi = as.character(dat$CD127n.pct_CD8 > CD127n.pct_thr)
###dat$hi = as.character(dat$CD127n.pct_CD8 > median(dat$CD127n.pct_CD8))

#pval
print(survdiff(Surv(days, event) ~ hi, data = dat)$pvalue)
#hr
print(
  summary(coxph(Surv(days, event) ~ hi, data = dat))[["coefficients"]][1,'exp(coef)']
)

#pl
survfit(Surv(days, event) ~ hi, data = dat) %>% 
  ggsurvfit(type = "risk", xscale='d_m')+
  add_confidence_interval()+
  add_censor_mark(size=1)+
  scale_ggsurvfit(x_scales = list( 
    breaks = seq(0,365*3.5,365/2), labels=seq(0,42,6), limits=c(0,365*3) 
  ))+
  add_risktable()+
  scale_y_continuous(breaks=seq(0,1,0.2),labels=seq(0,100,20),limits=c(0,1))+
  scale_color_manual(values = c('grey50','#eb7a09'))+
  scale_fill_manual(values = c('grey50','#eb7a09'))+
  labs(x='Months',y='% progressed')+
  theme_classic()+
  theme(legend.position='bottom')->pl
pl
ggsave(plot=print(pl, newpage = FALSE),filename = '../figures/flow/surv-CD127n.pdf',width=3.5,height=4, dpi="retina")

```


CD3+
```{r}
dat = import %>% 
  filter(timepoint=='Baseline',!is.na(event), !is.na(CD127p.pct_CD3))

dat$hi = dat$CD127p.pct_CD3 > median(dat$CD127p.pct_CD3)
print(survdiff(Surv(days, event) ~ hi, data = dat)$pvalue)

```


## CyTOF Treg SMM/MM

```{r fig.width=1.25, fig.height=3} 

dat = read_csv('/Volumes/ritd-ag-project-rd017s-klyon56/Atlas_analysis_code/R-resub/data/CyTOF/CyTOF_reGateCGK-Proportions.csv') %>% left_join(read_csv('/Volumes/ritd-ag-project-rd017s-klyon56/Atlas_analysis_code/R-resub/data/CyTOF/CyTOF_reGateCGK-md.csv'))

dat %>% mutate(cohort=factor(cohort,c('SMM','MM'))) %>% 
  ggplot(aes(cohort,FOXP3p_CD25p_pctCD4,fill=cohort))+
  
  geom_boxplot(outlier.alpha = 0)+
  geom_jitter(width=0.2,height=0,size=0.2)+
  ggpubr::stat_compare_means(comparisons = list(c('SMM','MM')), method = 'wilcox.test',size=3.5)+
  scale_fill_manual(values=diagnosis_col)+
  #scale_y_log10()+
  #scale_y_continuous(breaks=c(0,0.25,0.5),limits=c(0,0.75),labels=c('0','0.25','0.5'))+
  labs(title='')+

  theme_classic()+
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size=12),
    axis.title.x = element_blank(), 
    axis.text.x = element_text(angle = 45, vjust = 1, hjust=1, size=11),
    axis.title.y = element_text(size=13), 
    axis.text.y = element_text(size=11)
  )->pl
pl
ggsave(plot=pl,filename = '../figures/flow/box-CyTOF_Treg.pdf',width=1.25,height=3, dpi="retina")

```


## flow_Treg vs atlas Treg pseudo-pct non-tumour

```{r fig.height=2.5, fig.width=3.5}

dat = comp$Tcell %>% filter(celltype=='CD4+Treg') %>% 
  filter( ! sample_id %in% sort_id[sort_id$sort_id!='T cell-enriched',]$sample_id ) %>% 
  select(donor_id,prop,cohort) %>% 
  left_join(read.csv('../resources/donor_id-flow_scRNA.csv')) %>% drop_na() %>% 
  left_join(import %>% filter(timepoint=='Baseline') %>% 
              select(patient,treg.pct_nonTum,T.pct_nonTum)) %>% drop_na()

# Uses import$T.pct_nonTum

dat %>% 
  ggplot(aes(prop*T.pct_nonTum, treg.pct_nonTum))+
  geom_point(aes(color=cohort))+scale_color_manual(values=diagnosis_col)+
  geom_smooth(method='lm',color='black')+stat_cor()+
  scale_y_log10(breaks=c(0.1,1))+scale_x_log10(breaks=c(0.1,1))+annotation_logticks(sides='lb')-> pl
pl
ggsave(plot=pl,filename = '../figures/flow/cor-scRNA_Treg_pctnon.pdf',width=3.5,height=2.5, dpi="retina")


```



## CyTOF two Treg definitions correlation

```{r fig.height=2.5, fig.width=3.5}
dat = read_csv('/Volumes/ritd-ag-project-rd017s-klyon56/Atlas_analysis_code/R-resub/data/CyTOF/CyTOF_reGateCGK-Proportions.csv') %>% left_join(read_csv('/Volumes/ritd-ag-project-rd017s-klyon56/Atlas_analysis_code/R-resub/data/CyTOF/CyTOF_reGateCGK-md.csv'))

dat %>% 
  ggplot(aes(FOXP3p_CD25p_pctCD4,CD217n_CD25p_pctCD4))+
  geom_point(aes(color=cohort))+scale_color_manual(values=diagnosis_col)+
  geom_smooth(method='lm',color='black')+stat_cor()->pl
pl
ggsave(plot=pl,filename = '../figures/flow/cor-CyTOF_Treg.pdf',width=3.5,height=2.5, dpi="retina")

```





## flow_Treg loss association atlas Treg subclusters

```{r}

dat = comp$Treg %>% 
  filter( ! sample_id %in% sort_id[sort_id$sort_id!='T cell-enriched',]$sample_id ) %>% 
  select(donor_id,celltype,prop,cohort) %>% 
  left_join(read.csv('../resources/donor_id-flow_scRNA.csv')) %>% drop_na() %>% 
  left_join(import %>% select(patient,treg.pct_nonTum,T.pct_nonTum)) %>% drop_na()

dat %>% 
  ggplot(aes(prop, treg.pct_nonTum))+
  facet_grid(~celltype)+
  geom_point(aes(color=cohort))+scale_color_manual(values=diagnosis_col)+
  geom_smooth(method='lm',color='black')+stat_cor()


```



## flow_Treg pct SMM prg/non

% CD4+
```{r fig.width=1.65, fig.height=3} 

dat = import %>% 
  filter(timepoint=='Baseline',!is.na(event), !is.na(treg.pct_CD4), days<(365*3))

dat$cohort_prg = dat$cohort
dat$cohort_prg[dat$event==1  ]='SMM_PRG'
dat$cohort_prg[dat$cohort_prg=='MGUS']

p0 = dat$treg.pct_CD4 %>% .[.>0] %>% min()

dat %>% filter(treg.pct_CD4<50) %>% 
  ggplot(aes(cohort_prg,treg.pct_CD4+p0,fill=cohort_prg))+
  
  geom_boxplot(outlier.alpha = 0)+
  geom_jitter(width=0.2,height=0,size=0.2)+
  ggpubr::stat_compare_means(comparisons =
    list(c('MGUS','SMM'),c('MGUS','SMM_PRG'),c('SMM','SMM_PRG')),
  method = 'wilcox.test',size=3.5)+
  scale_fill_manual(values=diagnosis_col)+
  #scale_y_log10()+annotation_logticks(sides='l')+
  #scale_y_continuous(breaks=c(0,0.25,0.5),limits=c(0,0.75),labels=c('0','0.25','0.5'))+
  geom_hline(yintercept = treg.pctCD4_thr, color='blue')+
  labs(y='CD4+Treg % CD4+',title='')+
  theme_classic()+
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size=12),
    axis.title.x = element_blank(), 
    axis.text.x = element_text(angle = 45, vjust = 1, hjust=1, size=11),
    axis.title.y = element_text(size=13), 
    axis.text.y = element_text(size=11)
  )->pl
pl
ggsave(plot=pl,filename = '../figures/flow/box-flow_Treg_pctCD4.pdf',width=1.65,height=3, dpi="retina")


```


% non-tumour
```{r fig.width=1.8, fig.height=3} 

dat = import %>% 
  filter(timepoint=='Baseline',!is.na(event), !is.na(treg.pct_nonTum), days<(365*3))

dat$cohort_prg = dat$cohort
dat$cohort_prg[dat$event==1  ]='SMM_PRG'
dat$cohort_prg[dat$cohort_prg=='MGUS']

p0 = dat$treg.pct_nonTum %>% .[.>0] %>% min()

dat %>% filter(treg.pct_nonTum<5) %>% 
  ggplot(aes(cohort_prg,treg.pct_nonTum+p0,fill=cohort_prg))+
  
  geom_boxplot(outlier.alpha = 0)+
  geom_jitter(width=0.2,height=0,size=0.2)+
  ggpubr::stat_compare_means(comparisons =
    list(c('MGUS','SMM'),c('MGUS','SMM_PRG'),c('SMM','SMM_PRG')),
  method = 'wilcox.test',size=3.5)+
  scale_fill_manual(values=diagnosis_col)+
  scale_y_log10()+annotation_logticks(sides='l')+
  #scale_y_continuous(breaks=c(0,0.25,0.5),limits=c(0,0.75),labels=c('0','0.25','0.5'))+
  labs(y='CD4+Treg % non-tumour',title='')+
  geom_hline(yintercept = treg.pctNon_thr, color='blue')+

  theme_classic()+
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size=12),
    axis.title.x = element_blank(), 
    axis.text.x = element_text(angle = 45, vjust = 1, hjust=1, size=11),
    axis.title.y = element_text(size=13), 
    axis.text.y = element_text(size=11)
  )->pl
pl
ggsave(plot=pl,filename = '../figures/flow/box-flow_Treg_pctNon.pdf',width=1.8,height=3, dpi="retina")

```


## Supp: Treg vs individual risk metrics

% CD4+
```{r fig.height=2.5, fig.width=3.5}

dat = import %>% 
  filter(timepoint=='Baseline',!is.na(Entry_LB_paraprotein),!is.na(treg.pct_CD4)) 
dat %>% 
  ggplot(aes(Entry_LB_paraprotein,treg.pct_CD4))+
  geom_point(aes(color=cohort))+scale_color_manual(values=diagnosis_col)+
  geom_smooth(method='lm',color='black')+stat_cor()-> pl
pl
ggsave(plot=pl,filename = '../figures/flow/cor-treg.pct_CD4.PP.pdf',width=3.5,height=2.5, dpi="retina")

# BM PC %
dat = import %>% 
  filter(timepoint=='Baseline',!is.na(tumour.pct_live),!is.na(treg.pct_CD4),treg.pct_CD4<50) 
dat %>% 
  ggplot(aes(tumour.pct_live,treg.pct_CD4))+
  geom_point(aes(color=cohort))+scale_color_manual(values=diagnosis_col)+
  geom_smooth(method='lm',color='black')+stat_cor()+
  scale_x_log10()+annotation_logticks(sides='b')-> pl
pl
ggsave(plot=pl,filename = '../figures/flow/cor-treg.pct_CD4.BM_PC.pdf',width=3.5,height=2.5, dpi="retina")

dat = import %>% 
  filter(timepoint=='Baseline',!is.na(Entry_LB_FLC_ratio),!is.na(treg.pct_CD4),treg.pct_CD4<50) 
p0=min(dat$Entry_LB_FLC_ratio %>% .[.>0])
dat %>% 
  ggplot(aes(Entry_LB_FLC_ratio+p0,treg.pct_CD4))+
  geom_point(aes(color=cohort))+scale_color_manual(values=diagnosis_col)+
  geom_smooth(method='lm',color='black')+stat_cor()+
  scale_x_log10(breaks=c(0.01,0.1,1,10,100),labels=c(0.01,0.1,1,10,100))+
  annotation_logticks(sides='b')
#ggsave(plot=pl,filename = '../figures/flow/cor-CD127.PP.pdf',width=3.5,height=2.5, dpi="retina")


```



% non-tumour
```{r fig.height=2.5, fig.width=3.5}

dat = import %>% 
  filter(timepoint=='Baseline',!is.na(Entry_LB_paraprotein),!is.na(treg.pct_nonTum)) 
dat %>% 
  ggplot(aes(Entry_LB_paraprotein,treg.pct_nonTum))+
  geom_point(aes(color=cohort))+scale_color_manual(values=diagnosis_col)+
  geom_smooth(method='lm',color='black')+stat_cor()-> pl
pl
ggsave(plot=pl,filename = '../figures/flow/cor-treg.pct_nonTum.PP.pdf',width=3.5,height=2.5, dpi="retina")

# BM PC %
dat = import %>% 
  filter(timepoint=='Baseline',!is.na(tumour.pct_live),!is.na(treg.pct_nonTum),treg.pct_nonTum<5) 
dat %>% 
  ggplot(aes(tumour.pct_live,treg.pct_nonTum))+
  geom_point(aes(color=cohort))+scale_color_manual(values=diagnosis_col)+
  geom_smooth(method='lm',color='black')+stat_cor()+
  scale_x_log10()+annotation_logticks(sides='b')-> pl
pl
ggsave(plot=pl,filename = '../figures/flow/cor-tumour.pct_live.BM_PC.pdf',width=3.5,height=2.5, dpi="retina")

dat = import %>% 
  filter(timepoint=='Baseline',!is.na(Entry_LB_FLC_ratio),!is.na(treg.pct_nonTum),treg.pct_nonTum<5) 
p0=min(dat$Entry_LB_FLC_ratio %>% .[.>0])
dat %>% 
  ggplot(aes(Entry_LB_FLC_ratio+p0,treg.pct_nonTum))+
  geom_point(aes(color=cohort))+scale_color_manual(values=diagnosis_col)+
  geom_smooth(method='lm',color='black')+stat_cor()+
  scale_x_log10(breaks=c(0.01,0.1,1,10,100),labels=c(0.01,0.1,1,10,100))+
  annotation_logticks(sides='b')
#ggsave(plot=pl,filename = '../figures/flow/cor-CD127.PP.pdf',width=3.5,height=2.5, dpi="retina")


```




## flow_Treg TTP

% CD4+
```{r fig.height=4, fig.width=3.5}

dat = import %>% 
  filter(timepoint=='Baseline',!is.na(event), !is.na(treg.pct_CD4), days < 3*365)

dat$hi = as.character(dat$treg.pct_CD4 < treg.pctCD4_thr)

#pval
print(survdiff(Surv(days, event) ~ hi, data = dat)$pvalue)

#hr
print(
  summary(coxph(Surv(days, event) ~ hi, data = dat))[["coefficients"]][1,'exp(coef)']
)

#pl
survfit(Surv(days, event) ~ hi, data = dat) %>% 
  ggsurvfit(type = "risk", xscale='d_m')+
  add_confidence_interval()+
  add_censor_mark(size=1)+
  scale_ggsurvfit(x_scales = list( 
    breaks = seq(0,365*3.5,365/2), labels=seq(0,42,6), limits=c(0,365*3) 
  ))+
  add_risktable()+
  scale_y_continuous(breaks=seq(0,1,0.2),labels=seq(0,100,20),limits=c(0,1))+
  scale_color_manual(values = c('grey50','#eb7a09'))+
  scale_fill_manual(values = c('grey50','#eb7a09'))+
  labs(x='Months',y='% progressed')+
  theme_classic()+
  theme(legend.position='bottom')->pl
pl
ggsave(plot=print(pl, newpage = FALSE),filename = '../figures/flow/surv-Treg_pctCD4.pdf',width=3.5,height=4, dpi="retina")


```

% non-tumour
```{r fig.height=4, fig.width=3.5}

dat = import %>% 
  filter(timepoint=='Baseline',!is.na(event), !is.na(treg.pct_nonTum), days < 3*365)

#dat$hi = dat$treg.pct_nonTum > median(dat$treg.pct_nonTum)
dat$hi = as.character(dat$treg.pct_nonTum < treg.pctNon_thr)

#pval
print(survdiff(Surv(days, event) ~ hi, data = dat)$pvalue)

#hr
print(
  summary(coxph(Surv(days, event) ~ hi, data = dat))[["coefficients"]][1,'exp(coef)']
)

#pl
survfit(Surv(days, event) ~ hi, data = dat) %>% 
  ggsurvfit(type = "risk", xscale='d_m')+
  add_confidence_interval()+
  add_censor_mark(size=1)+
  scale_ggsurvfit(x_scales = list( 
    breaks = seq(0,365*3.5,365/2), labels=seq(0,42,6), limits=c(0,365*3) 
  ))+
  add_risktable()+
  scale_y_continuous(breaks=seq(0,1,0.2),labels=seq(0,100,20),limits=c(0,1))+
  scale_color_manual(values = c('grey50','#eb7a09'))+
  scale_fill_manual(values = c('grey50','#eb7a09'))+
  labs(x='Months',y='% progressed')+
  theme_classic()+
  theme(legend.position='bottom')->pl
pl
ggsave(plot=print(pl, newpage = FALSE),filename = '../figures/flow/surv-Treg_pctNon.pdf',width=3.5,height=4, dpi="retina")


```

## flow_Treg TPP, no MGUS stats

```{r}

print('treg.pct_CD4')
dat = import %>% 
  filter(timepoint=='Baseline',!is.na(event), !is.na(treg.pct_CD4), days < 3*365, cohort=='SMM')
dat$hi = as.character(dat$treg.pct_CD4 < treg.pctCD4_thr)
print(survdiff(Surv(days, event) ~ hi, data = dat)$pvalue)
print(
  summary(coxph(Surv(days, event) ~ hi, data = dat))[["coefficients"]][1,'exp(coef)']
)

print('treg.pct_nonTum')
dat = import %>% 
  filter(timepoint=='Baseline',!is.na(event), !is.na(treg.pct_nonTum), days < 3*365, cohort=='SMM')
dat$hi = as.character(dat$treg.pct_nonTum < treg.pctNon_thr)
print(survdiff(Surv(days, event) ~ hi, data = dat)$pvalue)
print(
  summary(coxph(Surv(days, event) ~ hi, data = dat))[["coefficients"]][1,'exp(coef)']
)


```


## Treg TPP multivariate: individual risk features

### % non-tumour

```{r fig.height=1.75, fig.width=4.5}
#tumour.pct_live
#Entry_LB_paraprotein
#Entry_LB_FLC_ratio

dat = import %>% 
  filter(timepoint=='Baseline',!is.na(event),days < 3*365,
         !is.na(treg.pct_nonTum), !is.na(tumour.pct_live)) 
dat$Treg_ = factor(ifelse(dat$treg.pct_nonTum > treg.pctNon_thr,'hi','lo'),c('hi','lo'))
dat$tumour_ = factor(ifelse(dat$tumour.pct_live > median(dat$tumour.pct_live),'hi','lo'),c('lo','hi'))
cox_model = coxph(Surv(days, event) ~ Treg_ + tumour_, data = dat)
summary(cox_model)$coefficients[1,5] #0.03


dat = import %>% 
  filter(timepoint=='Baseline',!is.na(event),days < 3*365,
         !is.na(treg.pct_nonTum), !is.na(Entry_LB_paraprotein)) 
dat$Treg_ = factor(ifelse(dat$treg.pct_nonTum > treg.pctNon_thr,'hi','lo'),c('hi','lo'))
dat$tumour_ = factor(ifelse(dat$Entry_LB_paraprotein > median(dat$Entry_LB_paraprotein),'hi','lo'),c('lo','hi'))
cox_model = coxph(Surv(days, event) ~ Treg_ + tumour_, data = dat)
summary(cox_model)$coefficients[1,5] #0.1

dat = import %>% 
  filter(timepoint=='Baseline',!is.na(event),days < 3*365,
         !is.na(treg.pct_nonTum), !is.na(Entry_LB_FLC_ratio)) 
dat$Treg_ = factor(ifelse(dat$treg.pct_nonTum > treg.pctNon_thr,'hi','lo'),c('hi','lo'))
dat$tumour_ = factor(ifelse(dat$Entry_LB_FLC_ratio > median(dat$Entry_LB_FLC_ratio),'hi','lo'),c('lo','hi'))
cox_model = coxph(Surv(days, event) ~ Treg_ + tumour_, data = dat)
summary(cox_model)$coefficients[1,5] #***


dat = import %>% 
  filter(timepoint=='Baseline',!is.na(event),days < 3*365,
         !is.na(treg.pct_nonTum), 
         !is.na(tumour.pct_live), !is.na(Entry_LB_paraprotein), !is.na(Entry_LB_FLC_ratio)) 
dat$Treg_ = factor(ifelse(dat$treg.pct_nonTum > treg.pctNon_thr,'hi','lo'),c('hi','lo'))

#Treg lo and cont
cox_model = coxph(Surv(days, event) ~ Treg_ + tumour.pct_live + Entry_LB_paraprotein + Entry_LB_FLC_ratio, data = dat)
summary(cox_model)$coefficients[1,5]
#cat
dat$BM_PC_ = factor(ifelse(dat$tumour.pct_live > median(dat$tumour.pct_live),'hi','lo'),c('lo','hi'))
dat$PP_ = factor(ifelse(dat$Entry_LB_paraprotein > median(dat$Entry_LB_paraprotein),'hi','lo'),c('lo','hi'))
dat$FLCr_ = factor(ifelse(dat$Entry_LB_FLC_ratio > median(dat$Entry_LB_FLC_ratio),'hi','lo'),c('lo','hi'))
cox_model = coxph(Surv(days, event) ~ Treg_ + BM_PC_ + PP_ + FLCr_, data = dat)
summary(cox_model)$coefficients[1,5]

```


```{r fig.height=1.75, fig.width=4.5}
cox_summary=summary(cox_model)
#Extract hazard ratios, confidence intervals, and p-values
hazard_ratios <- exp(coef(cox_model))
conf_int <- exp(confint(cox_model))
p_values <- cox_summary$coefficients[,5]
#Plot
dat.pl = data.frame(
  Var1 = rownames(cox_summary$conf.int),
  pvals=p_values,
  hr = hazard_ratios,
  lower = conf_int[,1],
  upper = conf_int[,2]
)
dat.pl = dat.pl %>% data.frame(Var1=dat.pl$Var1,
  med=c(median(dat$treg.pct_nonTum),median(dat$tumour.pct_live),
        median(dat$Entry_LB_paraprotein),median(dat$Entry_LB_FLC_ratio)))

dat.pl %>% 
  
  left_join(
    dat %>% select(Treg_,BM_PC_,PP_,FLCr_) %>% 
      pivot_longer(everything()) %>% group_by_all() %>% tally() %>% 
      pivot_wider(names_from=value,values_from=n) %>% 
      mutate(name=paste0(name,'hi'),name=ifelse(name=='Treg_hi','Treg_lo',name)),
    by=c('Var1'='name')
  ) %>% 
  mutate(Var1=paste0(Var1,', >',med,'\nhi=',hi,', lo=',lo)) %>% 
  arrange(-hr) %>% mutate(Var1=factor(Var1,Var1)) %>% 
  
  ggplot(aes(y = Var1, x = hr)) +
  geom_point(size = 3, color = "royalblue") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  geom_text(aes(label=round(pvals,3)),color='red')+
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10(breaks = c(0.1, 1, 2, 5, 10)) +
  labs(
    #title = paste0(
    #  'vs MGUS n=',table(dat$Entry_Mayo_)[1],', ',
    #  'vs Treg_lo n=',table(dat$Treg_)['hi']
    #),
    x = "Hazard ratio (95% CI)",
    y = "Variable"
  )+
  theme(
    plot.title = element_text(hjust = 0.5, size=8),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12)
  )->pl
pl
ggsave(plot=pl,filename = '../figures/flow/forest-TregNon_metrics.pdf',width=4.5,height=1.75, dpi="retina")
```

## Treg TPP multivariate: risk groups

### % non-tumour

Mayo
```{r fig.height=1.75, fig.width=4.5}

dat = import %>% 
  filter(timepoint=='Baseline',!is.na(event),
         !is.na(treg.pct_nonTum), days < 3*365) %>%
  mutate(Entry_Mayo_=ifelse(Entry_Mayo %in% c('high','UHR'),'high/UHR',Entry_Mayo)) %>% 
  mutate(Entry_Mayo_=factor(Entry_Mayo_,c('MGUS','low','intermediate','high/UHR')))
  #with UHR
  #mutate(Entry_Mayo_=factor(Entry_Mayo,c('MGUS','low','intermediate','high','UHR'))) 

dat$Treg_ = factor(ifelse(dat$treg.pct_nonTum > treg.pctNon_thr,'hi','lo'),c('hi','lo'))

table(dat$Entry_Mayo_, dat$Treg_) #good mix

cox_model = coxph(Surv(days, event) ~ Entry_Mayo_ + Treg_, data = dat)
cox_summary <- summary(cox_model)
#Extract hazard ratios, confidence intervals, and p-values
hazard_ratios <- exp(coef(cox_model))
conf_int <- exp(confint(cox_model))
p_values <- cox_summary$coefficients[,5]
#Plot
dat.pl = data.frame(
  Var1 = rownames(cox_summary$conf.int),
  pvals=p_values,
  hr = hazard_ratios,
  lower = conf_int[,1],
  upper = conf_int[,2]
) %>% 
  left_join(data.frame(table(dat$Entry_Mayo_)) %>% mutate(Var1=paste0('Entry_Mayo_',Var1)) %>% 
              rbind(data.frame(Var1='Treg_lo',Freq=table(dat$Treg_)['lo'])))
dat.pl
dat.pl %>% 
  ggplot(aes(y = paste0(Var1,' n=',Freq), x = hr)) +
  geom_point(size = 3, color = "royalblue") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  geom_text(aes(label=round(pvals,3)),color='red')+
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10(breaks = c(0.1, 1, 10)) +
  labs(
    title = paste0(
      'vs MGUS n=',table(dat$Entry_Mayo_)[1],', ',
      'vs Treg_lo n=',table(dat$Treg_)['hi']
    ),
    x = "Hazard ratio (95% CI)",
    y = "Variable"
  )+
  theme(
    plot.title = element_text(hjust = 0.5, size=8),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12)
  )->pl
pl
ggsave(plot=pl,filename = '../figures/flow/forest-TregNon_Mayo.pdf',width=4.5,height=1.75, dpi="retina")

```

Entry_IMWG 20/2/20
```{r fig.height=1.75, fig.width=4.5}

dat = import %>% 
  filter(timepoint=='Baseline',!is.na(event),
         !is.na(treg.pct_nonTum), days < 3*365) %>%
  mutate(Entry_IMWG_=Entry_IMWG) %>% 
  mutate(Entry_IMWG_=ifelse(Entry_IMWG_ %in% c('high','UHR'),'high/UHR',Entry_IMWG_)) %>% 
  mutate(Entry_IMWG_=ifelse(Entry_IMWG_ %in% c('low','low-intermediate'),'low/low-int',Entry_IMWG_)) %>% 
  mutate(Entry_IMWG_=factor(Entry_IMWG_,c('MGUS','low/low-int','intermediate','high/UHR'))) 

dat$Treg_ = factor(ifelse(dat$treg.pct_nonTum > treg.pctNon_thr,'hi','lo'),c('hi','lo'))

#table(dat$Entry_IMWG_, dat$Treg_) good mix

cox_model = coxph(Surv(days, event) ~ Entry_IMWG_ + Treg_, data = dat)
cox_summary <- summary(cox_model)
#Extract hazard ratios, confidence intervals, and p-values
hazard_ratios <- exp(coef(cox_model))
conf_int <- exp(confint(cox_model))
p_values <- cox_summary$coefficients[,5]
#Plot
dat.pl = data.frame(
  Var1 = rownames(cox_summary$conf.int),
  pvals=p_values,
  hr = hazard_ratios,
  lower = conf_int[,1],
  upper = conf_int[,2]
) %>% 
  left_join(data.frame(table(dat$Entry_IMWG_)) %>% mutate(Var1=paste0('Entry_IMWG_',Var1)) %>% 
              rbind(data.frame(Var1='Treg_lo',Freq=table(dat$Treg_)['lo'])))
dat.pl
dat.pl %>% 
  ggplot(aes(y = paste0(Var1,' n=',Freq), x = hr)) +
  geom_point(size = 3, color = "royalblue") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  geom_text(aes(label=round(pvals,3)),color='red')+
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10(breaks = c(0.1, 1, 10)) +
  labs(
    title = paste0(
      'vs MGUS n=',table(dat$Entry_IMWG_)[1],', ',
      'vs Treg_lo n=',table(dat$Treg_)['hi']
    ),
    x = "Hazard ratio (95% CI)",
    y = "Variable"
  )+
  theme(
    plot.title = element_text(hjust = 0.5, size=8),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12)
  )->pl
pl
ggsave(plot=pl,filename = '../figures/flow/forest-TregNon_IMWG.pdf',width=4.5,height=1.75, dpi="retina")

```




### % CD4+

Mayo
```{r fig.height=1.75, fig.width=4.5}

dat = import %>% 
  filter(timepoint=='Baseline',!is.na(event),
         !is.na(treg.pct_CD4), days < 3*365) %>%
  mutate(Entry_Mayo_=ifelse(Entry_Mayo %in% c('high','UHR'),'high/UHR',Entry_Mayo)) %>% 
  mutate(Entry_Mayo_=factor(Entry_Mayo_,c('MGUS','low','intermediate','high/UHR')))
  #with UHR
  #mutate(Entry_Mayo_=factor(Entry_Mayo,c('MGUS','low','intermediate','high','UHR'))) 

dat$Treg_ = factor(ifelse(dat$treg.pct_CD4 > treg.pctCD4_thr,'hi','lo'),c('hi','lo'))

table(dat$Entry_Mayo_, dat$Treg_) #good mix

cox_model = coxph(Surv(days, event) ~ Entry_Mayo_ + Treg_, data = dat)
cox_summary <- summary(cox_model)
#Extract hazard ratios, confidence intervals, and p-values
hazard_ratios <- exp(coef(cox_model))
conf_int <- exp(confint(cox_model))
p_values <- cox_summary$coefficients[,5]
#Plot
dat.pl = data.frame(
  Var1 = rownames(cox_summary$conf.int),
  pvals=p_values,
  hr = hazard_ratios,
  lower = conf_int[,1],
  upper = conf_int[,2]
) %>% 
  left_join(data.frame(table(dat$Entry_Mayo_)) %>% mutate(Var1=paste0('Entry_Mayo_',Var1)) %>% 
              rbind(data.frame(Var1='Treg_lo',Freq=table(dat$Treg_)['lo'])))

dat.pl %>% 
  ggplot(aes(y = paste0(Var1,' n=',Freq), x = hr)) +
  geom_point(size = 3, color = "royalblue") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  geom_text(aes(label=round(pvals,3)),color='red')+
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10(breaks = c(0.1, 1, 10)) +
  labs(
    title = paste0(
      'vs MGUS n=',table(dat$Entry_Mayo_)[1],', ',
      'vs Treg_lo n=',table(dat$Treg_)['hi']
    ),
    x = "Hazard ratio (95% CI)",
    y = "Variable"
  )+
  theme(
    plot.title = element_text(hjust = 0.5, size=8),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12)
  )

```

Entry_IMWG 20/2/20
```{r fig.height=1.75, fig.width=4.5}

dat = import %>% 
  filter(timepoint=='Baseline',!is.na(event),
         !is.na(treg.pct_CD4), days < 3*365) %>%
  mutate(Entry_IMWG_=Entry_IMWG) %>% 
  mutate(Entry_IMWG_=ifelse(Entry_IMWG_ %in% c('high','UHR'),'high/UHR',Entry_IMWG_)) %>% 
  mutate(Entry_IMWG_=ifelse(Entry_IMWG_ %in% c('low','low-intermediate'),'low/low-int',Entry_IMWG_)) %>% 
  mutate(Entry_IMWG_=factor(Entry_IMWG_,c('MGUS','low/low-int','intermediate','high/UHR'))) 

dat$Treg_ = factor(ifelse(dat$treg.pct_CD4 > treg.pctCD4_thr,'hi','lo'),c('hi','lo'))

#table(dat$Entry_IMWG_, dat$Treg_) good mix

cox_model = coxph(Surv(days, event) ~ Entry_IMWG_ + Treg_, data = dat)
cox_summary <- summary(cox_model)
#Extract hazard ratios, confidence intervals, and p-values
hazard_ratios <- exp(coef(cox_model))
conf_int <- exp(confint(cox_model))
p_values <- cox_summary$coefficients[,5]
#Plot
dat.pl = data.frame(
  Var1 = rownames(cox_summary$conf.int),
  pvals=p_values,
  hr = hazard_ratios,
  lower = conf_int[,1],
  upper = conf_int[,2]
) %>% 
  left_join(data.frame(table(dat$Entry_IMWG_)) %>% mutate(Var1=paste0('Entry_IMWG__',Var1)) %>% 
              rbind(data.frame(Var1='Treg_lo',Freq=table(dat$Treg_)['lo'])))

dat.pl %>% 
  ggplot(aes(y = paste0(Var1,' n=',Freq), x = hr)) +
  geom_point(size = 3, color = "royalblue") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.2) +
  geom_text(aes(label=round(pvals,3)),color='red')+
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10(breaks = c(0.1, 1, 10)) +
  labs(
    title = paste0(
      'vs MGUS n=',table(dat$Entry_IMWG_)[1],', ',
      'vs Treg_lo n=',table(dat$Treg_)['hi']
    ),
    x = "Hazard ratio (95% CI)",
    y = "Variable"
  )+
  theme(
    plot.title = element_text(hjust = 0.5, size=8),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12)
  )

```



## Treg % non-tum vs atlas

```{r fig.height=2.5, fig.width=3.5}

dat = comp$Tcell %>% filter(celltype=='CD4+Treg') %>% 
  filter( ! sample_id %in% sort_id[sort_id$sort_id!='T cell-enriched',]$sample_id ) %>%
  select(sample_id,donor_id,prop,cohort) %>% 
  left_join(read.csv('../resources/donor_id-flow_scRNA.csv')) %>% drop_na() %>% 
  left_join(import %>% filter(timepoint=='Baseline') %>% 
              select(patient,treg.pct_nonTum,T.pct_nonTum)) %>% drop_na() %>% 
  left_join(ords$Tcell$PCA)

dat %>% 
  ggplot(aes(prop*T.pct_nonTum, PC1*-1))+
  geom_point(aes(color=cohort))+scale_color_manual(values=diagnosis_col)+
  geom_smooth(method='lm',color='black')+stat_cor()+
  scale_x_log10(breaks=c(0.1,1))+annotation_logticks(sides='lb')


dat = comp$Tcell %>% filter(celltype=='CD4+Treg') %>% 
  filter( ! sample_id %in% sort_id[sort_id$sort_id!='T cell-enriched',]$sample_id ) %>%
  select(sample_id,donor_id,prop,cohort) %>% 
  left_join(ords$Tcell$PCA)

dat %>% filter(cohort!='Non') %>% 
  ggplot(aes(prop, PC1*-1))+
  geom_point(aes(color=cohort))+scale_color_manual(values=diagnosis_col)+
  geom_smooth(method='lm',color='black')+stat_cor()+
  scale_x_log10(breaks=c(0.1,1))+annotation_logticks(sides='lb')


```

## Treg longitudinal (*cut*)

```{r fig.height=3, fig.width=5}

dat = import %>% 
  filter( patient %in% (import %>% filter(longit) %>% pull(patient)) ) %>% 
  mutate(prg = ifelse(patient %in% import[import$event==1,]$patient,'prg','non'))

dat %>% 
  ggplot(aes(days_from_last,treg.pct_nonTum))+
  facet_grid(~prg)+
  geom_line(aes(group=patient),linewidth=0.5,color='grey70')+geom_point()+
  geom_smooth(method='lm',se=T,fill='skyblue')+
  stat_cor(color='red')
  
dat %>% 
  ggplot(aes(days_from_bsl,treg.pct_nonTum))+
  facet_grid(~prg)+
  geom_line(aes(group=patient),linewidth=0.5,color='grey70')+geom_point()+
  geom_smooth(method='lm',se=T,fill='skyblue')+
  stat_cor(color='red')

```






## [*unused*] CyTOF TemTerm at baseline vs MRD 

```{r fig.width=1.25, fig.height=3} 

dat = read_csv('/Volumes/ritd-ag-project-rd017s-klyon56/Atlas_analysis_code/R-resub/data/CyTOF/CyTOF_reGateCGK-Proportions.csv') %>% left_join(read_csv('/Volumes/ritd-ag-project-rd017s-klyon56/Atlas_analysis_code/R-resub/data/CyTOF/CyTOF_reGateCGK-md.csv')) %>% select(-MRD)

dat = dat %>% left_join(
  read_csv('/Volumes/groupfolders/CI_Haematology_Yong/Kane Foster/manuscript2025-core_panel/radar_mrd_Jan25.csv') %>% mutate(patient=str_remove_all(patient,'-')),
  by=c('radar_real'='patient')
) %>% filter(!is.na(D100_MRD)) %>% 
  left_join(read.csv('~/Documents/ResearchProjects/RADAR_CyTOF/R/data/md.csv') %>% 
              select(patient_id,condition),by=c('radar_file'='patient_id'))

dat = dat %>% left_join(
  read_csv('/Volumes/ritd-ag-project-rd017s-klyon56/Atlas_analysis_code/R-resub/data/CyTOF/CyTOF-MRD_CTRU.csv'),by=c('radar_real'='patient_id')
)

#CD127n_pctCD8,Tem_pctCD8,Temra_pctCD8,CD57p_GZMBp_CD8

dat %>% 
  #mutate(cohort=factor(cohort,c('SMM','MM'))) %>% 
  ggplot(aes(MRD,CD57p_GZMBp_Tem+CD57n_GZMBp_Tem,fill=cohort))+
  
  geom_boxplot(outlier.alpha = 0)+
  geom_jitter(width=0.2,height=0,size=0.2)+
  ggpubr::stat_compare_means(comparisons = list(c('neg','pos')), 
                             method = 'wilcox.test',size=3.5)+
  scale_fill_manual(values=diagnosis_col)+
  #scale_y_log10()+
  #scale_y_continuous(breaks=c(0,0.25,0.5),limits=c(0,0.75),labels=c('0','0.25','0.5'))+
  labs(title='')+

  theme_classic()+
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size=12),
    axis.title.x = element_blank(), 
    axis.text.x = element_text(angle = 45, vjust = 1, hjust=1, size=11),
    axis.title.y = element_text(size=13), 
    axis.text.y = element_text(size=11)
  )
#->pl
#pl
#ggsave(plot=pl,filename = '../figures/flow/box-CyTOF_TemBM.pdf',width=1.25,height=3, dpi="retina")
```


## [*unused*] core panel MRD

```{r fig.height=3, fig.width=3}
x = read_csv('/Volumes/groupfolders/CI_Haematology_Yong/Kane Foster/manuscript2025-core_panel/all_combined_flow.csv') %>%
  #filter(study=='RADAR',timepoint %in% c('Diagnosis','Baseline')) %>% 
  filter(study=='RADAR',timepoint %in% c('100D','D100')) %>% 
  group_by(patient) %>% slice_head(n = 1) %>% ungroup()

x %>% left_join(
  read_csv('/Volumes/groupfolders/CI_Haematology_Yong/Kane Foster/manuscript2025-core_panel/radar_mrd_Jan25.csv')
) -> x

x %>% mutate(patient_id=str_remove_all(patient,'-')) %>%
  left_join(read_csv('/Volumes/ritd-ag-project-rd017s-klyon56/Atlas_analysis_code/R-resub/data/CyTOF/CyTOF-MRD_CTRU.csv')
) -> x

x = x %>% mutate(IL7Rneg_CD8 = 100-(CD38nCD127p.pct_CD8+CD38pCD127p.pct_CD8)) 

sum(!is.na(x$MRD))

#tumour.pct_live,CD57p.pct_CD8,CD127p.pct_CD8,treg.pct_nonTum
#CD38nCD127p.pct_CD8+CD38pCD127p.pct_CD8
#CD38pCD127n.pct_CD3/CD38nCD127p.pct_CD3

x %>% filter(!is.na(MRD)) %>% ggplot(aes(MRD,tumour.pct_live))+
  geom_boxplot(outlier.alpha=0)+
  geom_jitter(width=0.2,height=0,size=0.5)+ggpubr::stat_compare_means()+
  scale_y_log10()
#D100=***

x %>% filter(!is.na(MRD)) %>% ggplot(aes(MRD,CD8.pct_CD3))+
  geom_boxplot(outlier.alpha=0)+
  geom_jitter(width=0.2,height=0,size=0.5)+ggpubr::stat_compare_means()

x %>% filter(!is.na(MRD)) %>% ggplot(aes(MRD,CD57p.pct_CD8))+
  geom_boxplot(outlier.alpha=0)+
  geom_jitter(width=0.2,height=0,size=0.5)+ggpubr::stat_compare_means()

#for ( i in colnames(x)[6:52]){
#x$i=x[[i]]
#x %>% filter(!is.na(MRD)) %>% ggplot(aes(MRD,i))+
#  geom_boxplot(outlier.alpha=0)+ggtitle(i)+
#  geom_jitter(width=0.2,height=0,size=0.5)+ggpubr::stat_compare_means()->pl
#print(pl)}
```

age??
```{r}
x = read_csv('/Volumes/groupfolders/CI_Haematology_Yong/Kane Foster/manuscript2025-core_panel/all_combined_flow.csv') %>%
  filter(study=='RADAR',timepoint %in% c('Diagnosis','Baseline')) %>% 
  #filter(study=='RADAR',timepoint %in% c('100D','D100')) %>% 
  group_by(patient) %>% slice_head(n = 1) %>% ungroup()
x = x %>% mutate(IL7Rneg_CD8 = 100-(CD38nCD127p.pct_CD8+CD38pCD127p.pct_CD8)) 

x %>% left_join(
  read_csv('/Volumes/groupfolders/CI_Haematology_Yong/Kane Foster/manuscript2025-core_panel/all_clinical.csv') %>% select(patient,age) %>% distinct() %>% 
    group_by(patient) %>% slice_head(n = 1)
) %>% lm(CD8.pct_CD3 ~ MRD + age, .) %>% summary()


```

fold-change
```{r fig.height=3, fig.width=3}

x = read_csv('/Volumes/groupfolders/CI_Haematology_Yong/Kane Foster/manuscript2025-core_panel/all_combined_flow.csv') %>% 
  filter(study=='RADAR') %>% 
  filter(timepoint %in% c('Diagnosis','Baseline','100D','D100')) %>% 
  mutate(tp=ifelse(timepoint %in% c('100D','D100'),'tp2','tp1')) %>% 
  group_by(patient,tp) %>% slice_head(n = 1) %>% ungroup()

#tumour.pct_live,CD57p.pct_CD8,CD127p.pct_CD8,treg.pct_nonTum
#100-CD38nCD127p.pct_CD8+CD38pCD127p.pct_CD8
#CD38pCD127n.pct_CD3/CD38nCD127p.pct_CD3

x = x %>% mutate(IL7Rneg_CD8 = 100-(CD38nCD127p.pct_CD8+CD38pCD127p.pct_CD8)) 
#select(CD8.pct_CD3, CD8.pct_nonTum, IL7Rneg_CD8)
#CD8.pct_CD3 * IL7Rneg_CD8/100

x %>% 
  mutate(value= CD8.pct_CD3) %>% 
  select(patient,tp,value) %>% 
  pivot_wider(names_from=tp) %>% drop_na() %>% 
  mutate(fc=tp2/tp1, log2fc = log2(fc)) -> xx

xx %>% mutate(patient_id=str_remove_all(patient,'-')) %>%
  left_join(read_csv('/Volumes/ritd-ag-project-rd017s-klyon56/Atlas_analysis_code/R-resub/data/CyTOF/CyTOF-MRD_CTRU.csv')
) -> xx

xx %>% filter(!is.na(MRD)) %>% 
  ggplot(aes(MRD,log2fc))+
  geom_boxplot(outlier.alpha=0)+
  geom_jitter(width=0.2,height=0,size=0.1)+ggpubr::stat_compare_means()

xx %>% ggplot(aes(tp1,tp2))+geom_point(size=0.5)+geom_smooth(method='lm')

xx %>% filter(!is.na(MRD)) %>% 
  ggplot(aes(tp1,tp2,color=MRD))+geom_point(size=0.5)+geom_smooth(method='lm',se=F)

```

improving
```{r fig.height=3, fig.width=3}

x = read_csv('/Volumes/groupfolders/CI_Haematology_Yong/Kane Foster/manuscript2025-core_panel/all_combined_flow.csv') %>%
  #filter(study=='RADAR',timepoint %in% c('Diagnosis','Baseline')) %>% 
  filter(study=='RADAR',timepoint %in% c('100D','D100')) 
  
x = x %>% filter(patient %in% (table(x$patient) %>% .[.<2] %>% names()))

x = x %>% 
  group_by(patient) %>% slice_head(n = 1) %>% ungroup() %>% 
  mutate(patient_id=str_remove_all(patient,'-')) %>%
    left_join(read_csv('/Volumes/ritd-ag-project-rd017s-klyon56/Atlas_analysis_code/R-resub/data/CyTOF/CyTOF-MRD_CTRU.csv')
    ) %>% 
  mutate(IL7Rneg_CD8 = 100-(CD38nCD127p.pct_CD8+CD38pCD127p.pct_CD8)) %>% 
  left_join(
    read_csv('/Volumes/groupfolders/CI_Haematology_Yong/Kane Foster/manuscript2025-core_panel/all_clinical.csv') %>% select(patient,age) %>% distinct() %>% 
    group_by(patient) %>% slice_head(n = 1)
)

x %>% filter(!is.na(MRD)) %>% 
  ggplot(aes(MRD,treg.pct_nonTum))+
  geom_boxplot(outlier.alpha=0)+
  geom_jitter(width=0.2,height=0,size=0.5)+ggpubr::stat_compare_means()

#summary(lm(CD8.pct_CD3 ~ MRD + age, x))

#only want to remove are
#after certain date
#high risk recieved ISA
#before, had no

#RiskCTRUF50	Disease Risk CTRU	RISKCTRUF50F.	"1 = Standard-risk, 2 = High-risk, 3 = Not viable, 4 = Unknown, 5 = Unable to determine, 999 = Other, 9876 = Missing"
pt_stnd_risk = read_csv('/Volumes/groupfolders/CI_Haematology_Yong/Kane Foster/manuscript2025-core_panel/radar-risk.csv') %>% filter(RiskCTRUF50==1) %>% pull(patient)

x %>% filter(!is.na(MRD)) %>% 
  filter(patient_id %in% pt_stnd_risk) %>% 
  ggplot(aes(MRD,CD8.pct_CD3))+
  geom_boxplot(outlier.alpha=0)+
  geom_jitter(width=0.2,height=0,size=0.5)+ggpubr::stat_compare_means()



```




## [*NK paper*] NK.CD56 TTP

assessment/thr
```{r}
#CD56_dim.pct_NK,CD56_br.pct_NK,CD56_dim.pct_nonTum,CD56_br.pct_nonTum
library(maxstat)

i='CD56_br.pct_nonTum'
dat = import %>% filter(timepoint=='Baseline', !is.na(event), days < 3*365)
dat$val = dat[[i]]
dat = dat[!is.na(dat$val),]
mtHL = maxstat.test(Surv(days, event) ~ val, data=dat, smethod="LogRank", pmethod="HL")
dat$hi = dat$val > mtHL$estimate
print(survdiff(Surv(days, event) ~ hi, data = dat)$pvalue)
print(summary(coxph(Surv(days, event) ~ hi, data = dat))[["coefficients"]][1,'exp(coef)'])

#val, p, hr
#CD56_dim.pct_NK 0.002423281, 0.3165103
#CD56_br.pct_NK 0.00242, 3.2
#CD56_dim.pct_nonTum, 0.002, 3.3
#CD56_br.pct_nonTum, 0.09, 2.2
```

surv curves
```{r fig.height=4, fig.width=3.5}
library(maxstat)

i='CD56_br.pct_NK'
i='CD56_dim.pct_NK'

dat = import %>% filter(timepoint=='Baseline', !is.na(event), days < 3*365)
dat$val = dat[[i]]
dat = dat[!is.na(dat$val),]
mtHL = maxstat.test(Surv(days, event) ~ val, data=dat, smethod="LogRank", pmethod="HL")
print(mtHL$estimate)
dat$hi = dat$val > mtHL$estimate

#p
print(survdiff(Surv(days, event) ~ hi, data = dat)$pvalue)
#hr
print(summary(coxph(Surv(days, event) ~ hi, data = dat))[["coefficients"]][1,'exp(coef)'])

#pl
survfit(Surv(days, event) ~ hi, data = dat) %>% 
  ggsurvfit(type = "risk", xscale='d_m')+
  add_confidence_interval()+
  add_censor_mark(size=1)+
  scale_ggsurvfit(x_scales = list( 
    breaks = seq(0,365*3.5,365/2), labels=seq(0,42,6), limits=c(0,365*3) 
  ))+
  add_risktable()+
  scale_y_continuous(breaks=seq(0,1,0.2),labels=seq(0,100,20),limits=c(0,1))+
  scale_color_manual(values = c('grey50','#eb7a09'))+
  scale_fill_manual(values = c('grey50','#eb7a09'))+
  labs(x='Months',y='% progressed')+
  theme_classic()+
  theme(legend.position='bottom')->pl
pl
ggsave(plot=print(pl, newpage = FALSE),
       filename = paste0('../figures/flow/surv-',i,'.pdf'),
       width=3.5,height=4, dpi="retina")


```


forest plot multivar risk
```{r fig.height=1.75, fig.width=4.5}

# mayo
i='CD56_br.pct_nonTum'
dat = import %>% filter(timepoint=='Baseline', !is.na(event), days < 3*365) %>% 
  mutate(Entry_Mayo_=ifelse(Entry_Mayo %in% c('high','UHR'),'high/UHR',Entry_Mayo)) %>% 
  mutate(Entry_Mayo_=factor(Entry_Mayo_,c('MGUS','low','intermediate','high/UHR')))
dat$val = dat[[i]]
dat = dat[!is.na(dat$val),]
mtHL = maxstat.test(Surv(days, event) ~ val, data=dat, smethod="LogRank", pmethod="HL")
dat$hi = dat$val > mtHL$estimate

dat$CD56_br_ = factor(ifelse(dat$hi,'hi','lo'),c('hi','lo'))

table(dat$Entry_Mayo_, dat$CD56_br_) #good mix

cox_model = coxph(Surv(days, event) ~ Entry_Mayo_ + CD56_br_, data = dat)
cox_summary <- summary(cox_model)

# imwg
i='CD56_br.pct_nonTum'
dat = import %>% filter(timepoint=='Baseline', !is.na(event), days < 3*365) %>% 
  mutate(Entry_IMWG_=Entry_IMWG) %>% 
  mutate(Entry_IMWG_=ifelse(Entry_IMWG_ %in% c('high','UHR'),'high/UHR',Entry_IMWG_)) %>% 
  mutate(Entry_IMWG_=ifelse(Entry_IMWG_ %in% c('low','low-intermediate'),'low/low-int',Entry_IMWG_)) %>% 
  mutate(Entry_IMWG_=factor(Entry_IMWG_,c('MGUS','low/low-int','intermediate','high/UHR'))) 
dat$val = dat[[i]]
dat = dat[!is.na(dat$val),]
mtHL = maxstat.test(Surv(days, event) ~ val, data=dat, smethod="LogRank", pmethod="HL")
dat$hi = dat$val > mtHL$estimate

dat$CD56_br_ = factor(ifelse(dat$hi,'hi','lo'),c('hi','lo'))

table(dat$Entry_IMWG_, dat$CD56_br_) #good mix

cox_model = coxph(Surv(days, event) ~ Entry_IMWG_ + CD56_br_, data = dat)
summary(cox_model)

```



## CyTOF and MRD

```{r fig.width=1.25, fig.height=3} 
dat = read_csv('/Users/kaneucl/Library/CloudStorage/OneDrive-Personal/COSMOS.Tcell/R-resub/data/CyTOF/CyTOF_reGateCGK-Proportions.csv') %>% left_join(read_csv('/Users/kaneucl/Library/CloudStorage/OneDrive-Personal/COSMOS.Tcell/R-resub/data/CyTOF/CyTOF_reGateCGK-md.csv')) %>% 
  left_join(read_csv('/Users/kaneucl/Library/CloudStorage/OneDrive-Personal/COSMOS.Tcell/R-resub/data/CyTOF/CyTOF-MRD_CTRU.csv'), by=c('radar_real'='patient_id'),suffix = c('.v1','.v2')) 

dat %>% filter(!is.na(MRD.v2)) %>% 
  mutate(cohort=factor(cohort,c('neg','pos'))) %>% 
  ggplot(aes(MRD.v2,Tem_pctCD8+Temra_pctCD8,fill=cohort))+
  
  geom_boxplot(outlier.alpha = 0)+
  geom_jitter(width=0.2,height=0,size=0.2)+
  ggpubr::stat_compare_means(comparisons = list(c('neg','pos')), method = 'wilcox.test',size=3.5)+
  scale_fill_manual(values=diagnosis_col)+
  labs(title='')+

  theme_classic()+
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size=12),
    axis.title.x = element_blank(), 
    axis.text.x = element_text(angle = 45, vjust = 1, hjust=1, size=11),
    axis.title.y = element_text(size=13), 
    axis.text.y = element_text(size=11)
  )
#->pl
#pl
###ggsave(plot=pl,filename = '../figures/flow/box-CyTOF_Treg.pdf',width=1.25,height=3, dpi="retina")
```


```{r fig.width=1.25, fig.height=3} 
for (i in colnames(dat)[2:22]){
  
dat %>% filter(!is.na(MRD.v2)) %>% 
  mutate(cohort=factor(cohort,c('neg','pos'))) %>% 
  ggplot(aes_string('MRD.v2',i,fill='cohort'))+
  
  geom_boxplot(outlier.alpha = 0)+
  geom_jitter(width=0.2,height=0,size=0.2)+
  ggpubr::stat_compare_means(comparisons = list(c('neg','pos')), method = 'wilcox.test',size=3.5)+
  scale_fill_manual(values=diagnosis_col)+
  labs(title=i)+

  theme_classic()+
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size=12),
    axis.title.x = element_blank(), 
    axis.text.x = element_text(angle = 45, vjust = 1, hjust=1, size=11),
    axis.title.y = element_text(size=13), 
    axis.text.y = element_text(size=11)
  )->pl
print(pl)
}
```

```{r fig.width=1.25, fig.height=3} 

dat$tmp = (dat$CD57p_GZMBn_Tem + dat$CD57p_GZMBp_Tem) / dat$Tem_pctCD8

dat %>% filter(!is.na(MRD.v2)) %>% 
  mutate(cohort=factor(cohort,c('neg','pos'))) %>% 
  ggplot(aes_string('MRD.v2','tmp',fill='cohort'))+
  
  geom_boxplot(outlier.alpha = 0)+
  geom_jitter(width=0.2,height=0,size=0.2)+
  ggpubr::stat_compare_means(comparisons = list(c('neg','pos')), method = 'wilcox.test',size=3.5)+
  scale_fill_manual(values=diagnosis_col)+
  labs(title=i)+

  theme_classic()+
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size=12),
    axis.title.x = element_blank(), 
    axis.text.x = element_text(angle = 45, vjust = 1, hjust=1, size=11),
    axis.title.y = element_text(size=13), 
    axis.text.y = element_text(size=11)
  )
```



# SPACE



































































