---
title: "S2"
output:
  html_document:
    code_folding: hide
    toc: true # table of content true
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  ## if you want number sections at each table header
params:
    args: myarg
---

# ...
## Setup

```{r setup-import, eval=T, hide=T}
source('../resources/aes.R')
da_md = read.csv('../preprocessing/data/import/metadata-donor.csv') %>% mutate(diagnosis=cohort)
# PC_scores
# PC_Ig
# PC_DEG
# PC_DEG + pw

dir.create('../figures/S2')
```

# [move, 1I WIP] 1

## 1G

```{r 1G, fig.height=1.5, fig.width=3, eval=T, hide=T}
m_id = read.csv('../preprocessing/data/mPC_id.csv')

dat = m_id %>% mutate(mPC=ifelse(mPC,'mPC','hPC')) %>% 
  group_by(donor_id,mPC) %>% tally() %>% ungroup() %>% 
  group_by(donor_id) %>% mutate(pct=n/sum(n)*100) %>% ungroup() %>% 
  left_join(da_md %>% select(donor_id,diagnosis) %>% distinct()) %>% 
  mutate(diagnosis=factor(diagnosis,c('MGUS','SMM','MM'))) %>% distinct() %>% 
  mutate(clonal=factor(ifelse(mPC=='mPC','Clonal','Non-clonal'),c('Non-clonal','Clonal')))
  
dat %>% 
  mutate(donor_id = factor(donor_id,arrange(dat[dat$mPC=='mPC',],pct)$donor_id)) %>% 
  ggplot(aes(donor_id,pct,fill=clonal))+
  geom_bar(stat='identity', width = 1)+
  scale_fill_manual(values = mPC_col)+
  labs(x='Patient (n = 46)',y='% plasma cells',fill='Immunoglobulin')+
  scale_y_continuous(limits=c(70,100), breaks = c(75,100),oob = scales::rescale_none)+
  theme(
    axis.text.x = element_blank(), axis.ticks.x = element_blank()
  ) -> pl
pl
#ggsave(plot=pl,filename = '../figures/1/1G.pdf',width=3,height=1.5, dpi="retina")


```


## 1H

```{r 1H, fig.width=5, fig.height=2, eval=T, hide=T}

load('../preprocessing/data/tumour_modules_scr.RData')
pathways = read.csv('resources/pathways.csv')
pathways_neat = pathways %>% select(-gene) %>% distinct()
scr_gavish = scr[1:40,]
scr_gavish$max = rownames(scr_gavish)[apply(assay(scr_gavish),2,FUN = which.max)]

dat = data.frame(table(scr_gavish$max)) %>% mutate(Var1=as.vector(Var1)) %>% arrange(-Freq) %>% 
  left_join(pathways_neat[1:40,] %>% mutate(Var1=rownames(scr)[1:40])) %>% 
  mutate(Var1=replace(pathway_neat, Freq < 2000 ,'Other')) %>% 
  group_by(Var1) %>% summarise(Freq=sum(Freq)) %>% 
  arrange(Freq) %>% mutate(Var1=factor(Var1,unique(Var1))) %>% 
  mutate(pct=round(Freq/sum(Freq)*100,2))
  
dat %>% 
  ggplot(aes(x="", y=Freq, fill=Var1))+
  geom_bar(stat="identity", size=0.5, color='black',linewidth=0.75)+
  coord_polar("y", start=0)+
  labs(fill='Dominant pathway')+
  scale_fill_brewer(palette = 'Set1',direction = -1, breaks=rev(dat$Var1))+
  #guides(fill=guide_legend(ncol=2))+
  theme_void()+
  theme(
    #plot.margin = margin(1,1,1,1, "cm"),
    legend.key.size = unit(0.5, "cm"),
    legend.text=element_text(size=10),
    legend.title=element_blank()
  ) -> pl
pl
#ggsave(plot=pl,filename = '../figures/1/1H.pdf',width=5,height=2, dpi="retina")

```



## 1I

comp.RData
PC_scores
_da_testing_

# S2

## Overview N donor/cells ig predicted

```{r Ncells, eval=T, hide=T}
m_id = read.csv('../preprocessing/data/mPC_id.csv')

m_id %>% filter(!donor_id %in% donors.longit) %>% dim()
m_id %>% filter(!donor_id %in% donors.longit) %>% pull(donor_id) %>% unique() %>% length()
pc_n = m_id %>% filter(!donor_id %in% donors.longit) %>% group_by(donor_id) %>% tally() %>% .$n 
median(pc_n)
range(pc_n)
```

## S2A

Data prep
```{r S2A-prep, eval=T, hide=T}

m_id = read.csv('../preprocessing/data/mPC_id.csv')

gene_n = m_id %>% select(donor_id,lightV1,lightV2,IGHV1,IGHV2,IGHC1,IGHC2) %>% 
  pivot_longer(!donor_id,names_to='gene') %>% filter(value!='') %>% 
  group_by(donor_id,gene,value) %>% tally() %>% ungroup() %>% 
  left_join(data.frame(gene=c('lightV1','lightV2','IGHV1','IGHV2','IGHC1','IGHC2'),
    group=c('light','light','IGHV','IGHV','IGHC','IGHC'),
    rank=c(1,2,1,2,1,2))) %>% 
  arrange(-n)

ig_gene_group = data.frame()
for (g in c('lightV','IGHC','IGHV')){
  gs=unique(c(m_id[[paste0(g,1)]],m_id[[paste0(g,1)]]))
  ig_gene_group = rbind(ig_gene_group,data.frame(gene=gs,group=rep(g,length(gs))) ) 
}
ig_gene_group = ig_gene_group %>% filter(gene!='')

m_id_pl = m_id %>% filter(!donor_id %in% donors.longit) %>% 
  select(donor_id,lightV1,IGHV1,IGHC1) %>% pivot_longer(!donor_id,names_to='chain',values_to='g') %>% 
  filter(g!='') %>% 
  group_by(donor_id,chain,g) %>% tally() %>% ungroup() %>% 
  group_by(donor_id,chain) %>% mutate(pct=n/sum(n)*100) %>% ungroup() %>% 
  mutate(x=paste0(donor_id,chain,g)) %>% arrange(pct) %>% mutate(x=factor(x,x)) %>% 
  arrange(-n) %>% group_by(donor_id,chain) %>% 
  mutate(rank=row_number()) %>% ungroup() %>% 
  mutate(rank=factor(ifelse(rank %in% 1:2,rank,'<2'),c(1,2,'<2'))) %>% 
  left_join(data.frame(
    chain=c("lightV1","IGHC1","IGHV1"),
    chain.pl=c('Light variable','Heavy constant','Heavy variable')
  )) %>% mutate(chain.pl=factor(chain.pl, c('Light variable','Heavy constant','Heavy variable')))

# Quantify rank 1 %s
tmp = m_id_pl %>% filter(rank==1) %>% group_by(chain.pl) %>% 
  summarise(med=median(pct), r1=range(pct)[1], r2=range(pct)[2])
print(tmp)
```

Plotting
```{r S2A, fig.height=2.5, fig.width=5, eval=T, hide=T}
donor.order = m_id_pl %>% filter(chain=='lightV1',rank==1) %>% arrange(-pct) %>% pull(donor_id)
m_id_pl_sub = m_id_pl %>% filter(donor_id==donor.order[27],rank!='<2',chain=='lightV1')

m_id_pl %>% 
  mutate(rank=ifelse(rank==1,1,'<1')) %>% 
  ggplot(aes(donor_id,pct))+facet_grid(~chain.pl)+
  geom_point(aes(color=rank),size=1.25)+
    scale_color_manual(values = setNames(c('#eb676e','grey77'),c(1,'<1')))+
  labs(x='Patient',y='% plasma cells\nexpressing gene',color='Gene rank')+
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        legend.position = 'bottom') -> pl
pl
#ggsave(plot=pl,filename = '../figures/S2/A.pdf',width=5,height=2.5, dpi="retina")
```


## Quantify most abundant Ig chains

- https://ashpublications.org/blood/article/110/9/3112/23967/Immunoglobulin-gene-rearrangements-and-the
  - VH3-30 (*IGHV*) has been consistently found to be preferentially rearranged in myeloma, followed by VH5-51, VH1-69, and VH3-23. H3-30 and VH3-23 are the most frequently used gene segments both in normal and malignant B-cell populations, not MM-specific
  - Most of the studies show a clear preference for Vκ1 (*IGKV*) family usage (53% of the total), followed by Vκ3 (25%); align w healthy B
  - a few Vκ1 (*IGKV*) gene segments seem to be used very often; these include the Vκ1-33/1D-33, Vκ1-39/1D-39, and Vκ1-6 gene segments
  - Studies dealing with Vλ (*IGLV*) usage in Igλ myeloma cases show largely equal Vλ1 (27%), Vλ2 (29%), and Vλ3 (41%) family usage (Table 4).46,59,77,78  Within these families, no clear preference for individual gene segments was observed.
  
```{r S2A-abundantIg, eval=T, hide=T}

m_id_pl %>% filter(rank==1,chain!='IGHC1') %>% 
  group_by(chain,g) %>% tally() %>% arrange(-n) %>% filter(n>1)

common.IGH = c('IGHV3-30',"IGHV5-51", 'IGHV1-69', 'IGHV3-23')
for (g in common.IGH ){print(paste0(g,'   ', g %in% unique(c(m_id$IGHV1,m_id$IGHV2))))}

common.lightV = c('IGKV1-33','IGKV1D-33','IGKV1-39','IGKV1D-39','IGKV1-6')
for (g in common.lightV ){print(paste0(g,'   ', g %in% unique(c(m_id$lightV1,m_id$lightV2))))}

m_id_pl %>% filter(rank==1,chain!='IGHC1') %>% group_by(chain,g) %>% 
  group_by(chain,g) %>% tally() %>% arrange(-n) %>% filter(n>1) %>% 
  mutate(is = g %in% c(common.IGH,common.lightV)) %>% filter(is) %>% 
  mutate(pct=n/length(unique(m_id_pl$donor_id))*100)

```

### Thresholding 


Overview


## S2B

PC_Ig
import/PC.h5ad


```{r S2B, fig.height=5, fig.width=12, eval=T, hide=T}
library(ComplexHeatmap)
library(circlize)

sce = zellkonverter::readH5AD('../preprocessing/data/import/PC.h5ad')
m_id = read.csv('../preprocessing/data/mPC_id.csv')

# Subset Ig genes
sce = sce[rownames(sce) %in% unique(c(m_id$lightV1,m_id$lightV2,m_id$IGHC1,m_id$IGHC2,m_id$IGHV1,m_id$IGHV2)),]
# Min expressed one cell
sce = sce[rowSums(assay(sce))>0,]

# Select donors
sce = sce[,sce$donor_id %in% m_id$donor_id]
sce = sce[,!sce$donor_id %in% donors.longit]

# Subset to mPC
sce = sce[,colnames(sce) %in% m_id$obs_names]

# Subset 50 cells per donor
index.plot = as_tibble(colData(sce)) %>% mutate(x=colnames(sce)) %>% 
  select(donor_id,x) %>% group_by(donor_id) %>% slice_head(n = 50) %>% pull(x)
sce=sce[,index.plot]

# Expresion
mat = as.matrix(assay(sce))
mat <- t(scale(t(mat)))

# Ig gene groups
genes=rownames(mat)
genes_gr = bind_rows(list(
  lightV=data.frame(gene=rownames(sce)[startsWith(rownames(sce),c('IGKV', 'IGLV'))]),
  IGHV=data.frame(gene=rownames(sce)[startsWith(rownames(sce),c('IGHV'))]),
  IGHC=data.frame(gene=c('IGHA2', 'IGHE', 'IGHG4', 'IGHG2', 'IGHGP', 'IGHA1', 'IGHG1', 'IGHG3', 'IGHD', 'IGHM'))
),.id='group') %>% 
  left_join(
    data.frame(group=c("lightV","IGHC","IGHV"),
               group_pl=c('Light\nvariable','Heavy\nconstant','Heavy\nvariable'))
  )
genes_gr = as_tibble(rowData(sce)) %>% left_join(genes_gr) %>% pull(group_pl)

# Donor col split
patient = sce$donor_id
donor_id_gr=sce$donor_id

#pl
hm_name='Scaled\nexpression'
titleSettings <- gpar(fontsize = 10)
textSettings_dims <- gpar(fontsize = 10)
  
hm <- Heatmap(mat, ##[,cell_order], 
  name = hm_name, column_title_side = 'top', row_title_side = 'left',
  col = colorRamp2(c(-0.5,0,1,3), cividis(4)),
  # Clustering data
  cluster_rows = F, cluster_columns = F,
  show_row_dend = F, show_column_dend = F, 
  show_row_names = F, # show gene names (usually yes)
  show_column_names = F, # !!! cell names, never enable!
  row_names_gp = textSettings_dims, column_names_gp = textSettings_dims,
  column_names_rot = 45, row_names_rot = 0,
  row_labels=genes,
  #Splitting rows/columns
  row_split = genes_gr, #factor(gene_groups, levels=unique(gene_groups)),
  column_split = donor_id_gr,
  row_gap = unit(0.5, "mm"),
  #cluster_column_slices=F,#this controls slice order!
  #ROW
  row_title_gp = titleSettings, # or set each individually
  row_title_rot = 0,  # only +/- units of 90 (vertical or horoz)
  #COL
  column_title = NULL,
  #column_title_gp = gpar(fontsize = 8), 
  #column_title_rot = 90, # only +/- units of 90 (vertical or horoz)
  # Annotation
  #top_annotation = column_ha,
  # Plotting
  use_raster=T,raster_quality=10
)

draw(hm)
#grDevices::png("../figures/S2/B.png", height = 350, width = 1000)
#ComplexHeatmap::draw(hm)
#grDevices::dev.off()
```



## S2C

- NSD2 = MMSET
- CCND1

```{r S2C, ffig.height=3.5, fig.width=6, eval=T, hide=T}

# Published translocation information
donor_transloc = data.frame(
  donor_id=c('Maura_2023.PT03','Maura_2023.PT12','Foster_2024.SMM8','Foster_2024.SMM10','Maura_2023.PT30','Maura_2023.PT63'),
  group=c('(4;14)+','(4;14)+','(4;14)+','(4;14)+','(11;14)+','(11;14)+'),
  donor_pl=c("Maura 2023\nPT03\nt(4;14)+","Maura 2023\nPT12\nt(4;14)+",'This study\nSMM-8\nt(4;14)+', 'This study\nSMM-10\nt(4;14)+',"Maura 2023\nPT30\nt(11;14)+","Maura 2023\nPT63\nt(11;14)+"))

# Expression
sce = zellkonverter::readH5AD('../preprocessing/data/import/PC.h5ad')
sce = sce[c('CCND1','NSD2'),sce$donor_id %in% donor_transloc$donor_id]

dat.pl = data.frame(donor_id=sce$donor_id, CCND1=as.vector(assay(sce['CCND1',])),
           MMSET=as.vector(assay(sce['NSD2',]))) %>% 
  mutate(mPC = colnames(sce) %in% m_id[m_id$mPC=='True',]$obs_names) %>% 
  mutate(mPC = ifelse(mPC,'Tumour (clonal)','Non-tumour (non-clonal)')) %>% 
  left_join(donor_transloc) %>% 
  select(donor_id,donor_pl,mPC,group,CCND1,MMSET) %>% 
  pivot_longer(!c(donor_id:group)) %>% 
  mutate(mPC=factor(mPC,levels=c('Non-tumour (non-clonal)','Tumour (clonal)'))) %>% 
  mutate(name=factor(name,c('MMSET','CCND1'))) %>% 
  mutate(donor_pl=factor(donor_pl,donor_transloc$donor_pl)) 

pl.col = c('Tumour (clonal)'='#CB6651','Non-tumour (non-clonal)'='#CBD5E8')  

dat.pl %>% 
  ggplot(aes(mPC,value))+facet_grid(name~donor_pl)+
  geom_violin(aes(fill=mPC))+geom_boxplot(alpha=0)+
  scale_fill_manual(values = pl.col)+
  stat_summary(fun.data = function(x){c(y = max(x)+0.5, label = length(x))}, 
    geom = "text", fun.y = median, size=3)+
  ggpubr::stat_compare_means(aes(label = paste0(after_stat(p.format))), 
    method='wilcox.test', label.y=5.75,label.x=1.25, size=3)+
  ylim(c(0,6.25))+
  labs(y='Expression', fill='')+
  theme(
    legend.position = 'bottom',
    legend.text = element_text(size=10),
    panel.border = element_rect(color = "black", fill = NA, size = 0.5),
    axis.title.x=element_blank(),axis.text.x = element_blank(),
    axis.text.y = element_text(size=11),
    strip.background=element_blank(),
    strip.text.x = element_text(size=9), strip.text.y=element_text(angle=0, hjust=1, size=11)
  )->pl
pl
#ggsave(plot=pl,filename = '../figures/S2/S2C.pdf',width=6,height=3.5, dpi="retina")

```


## S2D

```{r S2D, fig.height=3, fig.width=1.5, eval=T, hide=T}

m_id = read.csv('../preprocessing/data/mPC_id.csv')

dat = m_id %>% 
  filter(!donor_id %in% donors.longit) %>% 
  mutate(mPC=ifelse(mPC,'mPC','hPC')) %>% 
  group_by(donor_id,mPC) %>% tally() %>% ungroup() %>% 
  group_by(donor_id) %>% mutate(pct=n/sum(n)*100) %>% ungroup() %>% 
  left_join(da_md %>% select(donor_id,diagnosis) %>% distinct()) %>% 
  mutate(diagnosis=factor(diagnosis,c('MGUS','SMM','MM'))) %>% distinct()
  
dat %>% filter(mPC=='mPC') %>% 
  ggplot(aes(x = diagnosis, y = pct, fill=diagnosis))+
  geom_boxplot(outlier.alpha = 0)+
  geom_jitter(width=0.2,height=0,size=0.2)+
  ggpubr::stat_compare_means(comparisons = list(c('SMM','MM')),method = 'wilcox.test',size=3.5,label.y=102.5)+
  scale_y_continuous(limits = c(70,107.5), breaks = c(80,90,100))+
  scale_fill_manual(values=diagnosis_col)+
  labs(x='',y='% clonal plasma cells',title='')+
  theme_classic()+
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size=12),
    axis.title.x = element_blank(), 
    axis.text.x = element_text(angle = 45, vjust = 1, hjust=1, size=11),
    axis.title.y = element_text(size=11), 
    axis.text.y = element_text(size=11)
    )-> pl
pl
#ggsave(plot=pl,filename = '../figures/S2/S2D.pdf',width=1.5,height=3, dpi="retina")
```

## S2E

PC_scores

# [move, 5G WIP] 5

## 5F

Data prep
```{r 5F, fig.height=6, fig,width=7, eval=T, hide=T}
library(ComplexHeatmap)
tumour_DE = read.csv('../preprocessing/data/tumour_DE.csv',row.names=1)
load(file='../preprocessing/data/tumour_DE_fgsea.RData')

fgseas.select = fgseas[fgseas$pathway=='KEGG_ANTIGEN_PROCESSING_AND_PRESENTATION',] 
donor.sigAPP = fgseas.select[(fgseas.select$pval<0.05)&(fgseas.select$NES>0),]$donor_id

# Selected genes
KEGG_APC.g = c('HLA-A','HLA-B','HLA-C', 'B2M','HLA-DRA','HLA-DPA1','CD74','HLA-E','HLA-F','TAP1','TAP2','TAPBP','PSME1','PSME2')

donor.order = tumour_DE %>% filter(gene %in% KEGG_APC.g) %>% group_by(donor_id) %>%
  summarise(logFC=mean(logFC)) %>% arrange(logFC) %>% pull(donor_id)

dat.in = tumour_DE %>% filter(gene %in% KEGG_APC.g) %>%
    mutate(is = donor_id %in% donor.sigAPP) %>%
    mutate(donor_id = factor(donor_id,donor.order)) %>%
    group_by(gene) %>% mutate(logFC=as.vector(scale(logFC)))

dat.mat = dat.in %>% select(donor_id,gene,logFC) %>%
  mutate(donor_id=factor(donor_id,donor.order)) %>% 
  pivot_wider(names_from='donor_id',values_from='logFC',values_fill=NA) %>% 
  relocate(any_of(c('gene',donor.order))) %>% 
  mutate(gene=factor(gene,KEGG_APC.g)) %>% arrange(gene) %>% 
  data.frame() %>% column_to_rownames('gene') %>% as.matrix()

logFC.max=round(max(dat.in$logFC),2)
logFC.min=round(min(dat.in$logFC),2)
col_fun = circlize::colorRamp2(c(logFC.min, 0, logFC.max), c("dodgerblue", "white", "red"))

column_donor_md = dat.in %>% ungroup() %>% select(donor_id) %>% distinct() %>% 
  mutate(is_de = donor_id %in% donor.sigAPP) %>%
  left_join(
    read.csv('../preprocessing/data/import/metadata-donor.csv') %>% 
      select(donor_id,cohort) %>% distinct()
  ) %>% 
  mutate(donor_id=factor(donor_id,donor.order)) %>% arrange(donor_id) 

```

Plotting/saving
```{r fig.height=3, fig.width=4}

column_ha = columnAnnotation(
  " " = column_donor_md$cohort,
  col = list(` `=diagnosis_col),
  height = unit(4, "cm"),
  gp = gpar(col = "black"), gap = unit(0, "points"),
  annotation_name_side = "right", show_legend=F
)

row_ha = rowAnnotation(gene = anno_mark(at = 1:length(rownames(dat.mat)), labels = rownames(dat.mat)))

Heatmap(
  dat.mat,
  color=col_fun,
  na_col = 'grey80',
  top_annotation = column_ha,
  show_heatmap_legend=F,
  cluster_rows = F, cluster_columns = F,
  show_row_names = F, show_column_names = F, 
  right_annotation = row_ha, 
  row_names_gp = gpar(fontface='italic'),
  column_split = column_donor_md$is_de, border = TRUE, 
  column_title = c('MHC low', 'MHC high'),
  width = unit(6, "cm"), height = unit(4, "cm")
  ) -> hm

lgd_list = list(
  Legend(col_fun = col_fun, title = "Scaled log fold-change\ntumour / normal", at = c(logFC.min, 0, logFC.max), 
    direction = "horizontal",title_position = "topcenter",
    labels = c('Min', '', 'Max'),
    title_gp=gpar(fontface='plain')
  ))

draw(hm, annotation_legend_list = lgd_list, annotation_legend_side = "bottom")

#grDevices::pdf("../figures/5/F.pdf", width = 4, height = 3)
#draw(hm, annotation_legend_list = lgd_list, annotation_legend_side = "bottom")
#grDevices::dev.off()

```


## 5G

PC_DEG + pw
_nonviv_


# SPACE 








































































































